---
title: "Demographic Exploration and Discovery: Color in R"
author: "Tim Riffe"
date: "June 26, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Color

## Basics on specification

hex,
names,
ramps,
spaces

Let's talk about sequential, diverging, and qualitative colors
```{r}
library(colorspace)
hcl_palettes(plot = TRUE)
```
You can use these palettes like so to produce a vector of colors:
```{r}
sequential_hcl(10,"BurgYl")
qualitative_hcl(5,"Dynamic")
diverging_hcl(20, "Lisbon")

se <- sequential_hcl(10,"BurgYl")
demoplot(se,"map")
demoplot(se,"heatmap")
demoplot(se,"bar")
demoplot(se,"line")
demoplot(se,"scatter")
```

Or in the `ggplot2` framework:
```{r}
library("ggplot2")
ggplot(iris, aes(x = Sepal.Length, fill = Species)) + geom_density(alpha = 0.6) +
  scale_fill_discrete_qualitative(palette = "Dark 3")

dsamp <- diamonds[1 + 1:1000 * 50, ]
ggplot(dsamp, aes(carat, price, color = cut)) + geom_point() +
  scale_color_discrete_sequential(palette = "Purples 3", nmax = 6, order = 2:6)
```
You can compose the different `scale_color_` functions using the same pattern.

## Discrete vs continuous color

## Background vs foreground

## Points / lines vs areas

## geom_tile, geom_contour, geom_raster

```{r,fig.width=10,fig.height=5}
library(HMDHFDplus)
library(ggplot2)
library(colorspace)
DX <-readHMDweb(CNTRY = "SWE",
				item = "Deaths_1x1",
				username = us, 
				password = pw)


p <- ggplot(DX, mapping = aes(x = Year, y = Male, fill = Age)) 
	p + geom_col() + 
		scale_fill_continuous_sequential(palette = "BurgYl") 

p <- ggplot(DX, mapping = aes(x=Year, y=Age, fill = log(Male)))
p + geom_tile() +
	scale_fill_continuous_sequential(palette = "BurgYl") + 
	geom_contour(mapping=aes(z=log(Male)), col = "black") + 
	coord_fixed()

p + geom_raster() + coord_fixed()
```