---
title: "Class Notes"
author: "Tim Riffe"
date: "June 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Get packages loaded
```{r}
library(socviz)
library(gapminder)
library(tidyverse)
library(broom)
```
## List columns and nesting
We start off here: [https://socviz.co/modeling.html#grouped-analysis-and-list-columns](https://socviz.co/modeling.html#grouped-analysis-and-list-columns)

The objectives are to explain a column type that can hold complex data by using the `map` and `nest` functions. Since statistical model output is a complex list, we will use this column to store that kind of output, and we will send this to `ggplot` to make a nice plot of how the within-continent Preston relationship has changed over time.

```{r}

# this is a helper function.
fit_ols <- function(df){
	lm(lifeExp ~ log(gdpPercap), data = df)
}

out_le <- gapminder %>%
	group_by(continent, year) %>%
	nest() %>%
    mutate(model = map(data, fit_ols))

out_le
```

OK, so now we have a column called data that has the model output in each cell. But we can turn this into something useful and plottable!
Try again, using the tidy function
```{r}
out_tidy <- gapminder %>%
	group_by(continent, year) %>%
	nest() %>%
	mutate(model = map(data, fit_ols),
		   tidied = map(model, tidy)) %>%
	unnest(tidied, .drop = TRUE) %>%
	filter(term %nin% "(Intercept)" &
		   	continent %nin% "Oceania")
sample_n(out_tidy,5)
# Note %nin% is similar to !%in%
# take a look!
out_tidy %>% sample_n(5)
```

Great, now let's see how the slope has changed over time:
```{r}
p <- ggplot(data = out_tidy,
			mapping = aes(x = year,
						  y = estimate,
						  ymin = estimate - 2 * std.error,
						  ymax = estimate + 2 * std.error,
						  group = continent, color = continent))
p + geom_pointrange(position = position_dodge(width = 1.5)) + 
	scale_x_continuous(breaks = unique(gapminder$year))

```

Timeout for functions!

Toy example, arithmetic
```{r}
a <- function(x,y){
	b <- x+y * x^2 - y^3
	return(b - y + x)
}

a(x = runif(10), y = runif(10))
```

```{r}

mx2lx <- function(mx, radix){
	n <- length(mx)
	radix * c(1,exp(-cumsum(mx)))[1:n]
}

lx2dx <- function(lx){
	-diff(c(lx,0))
}

my_lt <- function(mx, radix){
	mx <- tibble(mx)
	lt <- mx %>% mutate(mx = mx,
				  lx = mx2lx(mx, radix = radix),
				  dx = lx2dx(lx))
	lt
}

```

# ggridges
Now we skip a couple chapters and jump to a ridge plot, because we'll probably want to make one on Thursday using demographic data.

```{r, fig.width = 5, fig.height = 10}
library(ggridges)
p <- ggplot(gss_lon,
			mapping = aes(x = age, 
						 y = factor(year, 
						 		   levels = rev(unique(year)), 
						 		   ordered = TRUE)))

p + geom_density_ridges(alpha = 0.6, fill = "lightblue", scale = 1.5) +
	scale_x_continuous(breaks = c(25,50,75)) + 
	scale_y_discrete(expand = c(0.01,0)) + 
	labs(x = "Age", y = NULL, title = "Age Distribution of \nGSS Respondents") + 
	theme_ridges() + 
	theme(title = element_text(size = 16, face = "bold"))
```

# Exercise for Thursday:

This is a challenge. Are you up to it?

Get HFC ASFR data from here:
[https://www.fertilitydata.org/cgi-bin/data.php](https://www.fertilitydata.org/cgi-bin/data.php)

Click on the one called *All HFC data (6.8Mb)* to get the big zip file.

Then follow these steps: 

1. move it to the folder for today's session
2. read it into `R` somehow.
3. examine the dataset using `sample_n()` to get a feel for it.
4. within each subset (there arte more grouping variables than just country and year) calculate the total fertility rate (TFR) and the (ASFR-weighted) mean age at childbearing (MAB).
5. now the goal is to draw a ridge plot of ASFR, where each ridge is a discrete bin of TFR in uniform intervals (maybe 8 - 15 different steps or so). Within each ridge there will be two distributions: The ASFR curve with the highest MAB and the one with the lowest MAB.
6. What is the message that this plot delivers?


