---
title: "Freestyle HFC ridgeplot"
author: "Tim Riffe"
date: "June 20, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Prelims:
```{r}
library(tidyverse)
library(ggplot2)
library(ggridges)
library(readr)
library(here)
```

# The exercise:
Get HFC ASFR data from here:
[https://www.fertilitydata.org/cgi-bin/data.php](https://www.fertilitydata.org/cgi-bin/data.php)

Click on the one called *All HFC data (6.8Mb)* to get the big zip file.

Then follow these steps: 

1. move it to the folder for today's session
2. read it into `R` somehow.

```{r}
HFC <- read_csv("HFC_ASFRstand_TOT.txt", 
				na = ".") # HMD world uses '.' for NA
# Warnings for the Note column. 
# No prob since we won't use that column
str(HFC)
table(HFC$Age) # uniform ages 14-50

```

3. examine the dataset using `sample_n()` to get a feel for it.

```{r}
sample_n(HFC,5)
```

4. within each subset (there arte more grouping variables than just country and year) calculate the total fertility rate (TFR) and the (ASFR-weighted) mean age at childbearing (MAB).

```{r}
# From the above, it seems like we have a lot of potential
# grouping variables
HFC <- HFC %>% 
	group_by(
	          Country, 
			  Year1,
			  Region, 
			  Urban, 
			  Origin, 
			  Collection, 
			  SourceType,
			  RefCode,
			  Vitality,
			  AgeDef) %>% 
	mutate(TFR = sum(ASFR, na.rm = TRUE),
		   MAB = sum(Age * ASFR, na.rm = TRUE) / TFR,
		   TFRint = TFR - TFR %% .25) %>% # can change TFR interval 
	ungroup() %>% 
	group_by(TFRint) %>% 
	mutate(MABmax = MAB == max(MAB),
		   MABmin = MAB == min(MAB),
		   MABminmax = MABmax | MABmin,
		   MABgroup = ifelse(MABmax, "max MAB",ifelse(MABmin,"min MAB",NA))) %>% 
	filter(MABminmax & TFR < 6) %>% 
	ungroup()
```

5. now the goal is to draw a ridge plot of ASFR, where each ridge is a discrete bin of TFR in uniform intervals (maybe 8 - 15 different steps or so). Within each ridge there will be two distributions: The ASFR curve with the highest MAB and the one with the lowest MAB.

```{r, fig.height = 10}
p <- ggplot(HFC,
			mapping = aes(x = Age, 
						 y = factor(TFRint, 
						 		   levels = sort(unique(TFRint)), 
						 		   ordered = TRUE)))
p + 
	geom_ridgeline(mapping = aes(x = Age, height = ASFR, fill = MABgroup),                          scale = 4, 
				   alpha = .6)
```
6. What is the message that this plot delivers?

