---
title: "Class Notes"
author: "Tim Riffe"
date: "July 10, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# variable recoding

Q: how to create a new variable with three categories based on two other variables.
```{r, message = FALSE}
library(tidyverse)
A <- data.frame(v1 = sample(c(1,2,9), 50, replace = TRUE),
				v2 = sample(c(1,2,9), 50, replace = TRUE)) %>% 
	mutate(v3 = case_when(v1 == 1 ~ 1,          # most general case
						  v1 == 2 & v2 == 1 ~ 2,
						  v1 == 9 & v2 == 9 ~ 9,
						  TRUE ~ 3, ))          # most specific case
A # just check this yourself
```


# More examples of pipelines

Here several `dplyr` verbs are showcased, but there are many many more. You can find a good overview here [https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

To reproduce this pipeline, you can get the data from this link:
[https://drive.google.com/open?id=1AibPQP8plaGv0MaRwOJqqOg2LR6wXqK0](https://drive.google.com/open?id=1AibPQP8plaGv0MaRwOJqqOg2LR6wXqK0)

```{r, message = FALSE}
library(tidyverse)
library(haven)
JG         <- read_spss("caseid_aggr.sav")

# in the original source.
activities <- attr(JG$time_1000_max, "label")

# reduce size so code runs faster while we test it.
# skip this line when ready for final graph
JG         <- sample_n(JG, 1000)

# pipeline starts here!
JG <- JG %>% 
	gather(key = time,
		   value = activity,
		   time_1_max:time_1440_max) %>% 
# now split the concatenated column called time
# keeping the middle bit and converting it to integer
	separate(col = time, 
			 into = c(NA, "time", NA),
			 sep = "_",
			 remove = TRUE,
			 convert = TRUE) %>% 
	filter(time %% 10 == 1) %>%  
	# select every 10th minute
	group_by(time, activity) %>%   
	summarize(n = n()) %>% 
	filter(time %% 10 == 1) %>%    # select every 10th minute
	group_by(time, activity) %>%   
	summarize(n = n()) %>%         # get frequencies
	mutate(prop = n / sum(n)) %>% 
	ungroup() %>% 
	mutate(activity = factor(
		activity,
		levels = activities,
		labels = names(activities)
	)) 
```

Running this pipeline from start to end transforms the data in discrete steps to a tidy dataset that is handy for plotting. Here was the desired plot:

```{r}
ggplot(JG, mapping = aes(x = time, y = prop, fill = activity)) +
	geom_area() +
	# tone down the colors
	scale_fill_discrete(c = 50, l = 80)
```