---
title: "My project"
author: "Tim Riffe"
date: "7/11/2019"
output: pdf_document
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

Recommended length: 100 - 150 words (doesn't count against words)
You can find a full description of the assignment here: [https://github.com/timriffe/DemoED/Projects.pdf](https://github.com/timriffe/DemoED/Projects.pdf)

# Background
Recommended length: ca 200 words
 Here is an awesome citation of an awesome paper that really puts this in perspective @lee2001evaluating


# Data and Methods
Recommended length: ca 100 words
```{r, results = "hide", echo = FALSE, message = FALSE, warning= FALSE}
# Read in the data

```

```{r, results = "hide", echo = FALSE, message = FALSE, warning= FALSE}
# prepare it for the model

```

```{r, results = "hide", echo = FALSE, message = FALSE, warning= FALSE}
# do the model

```

```{r, results = "hide", echo = FALSE, message = FALSE, warning= FALSE}
# tidy the output

```

# The visualization
Make sure it includes a healthy informative self-sufficient caption. Doesn't count against word limit.
```{r, echo = FALSE, message = FALSE, warning= FALSE}
```

# Discussion
recommended length 100-200 words

# References