---
title: "Setup"
author: "Tim Riffe"
date: "April 3, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup notes

First let's get some packages installed from CRAN. These are recommended by the Healy book. You should read Healy's essay "Fuck nuance".

```{r, eval = FALSE}
my_packages <- c("tidyverse","broom","coefplot","cowplot",
				 "gapminder","GGally","ggrepel","ggridges","gridExtra", "here", "interplot","margins","maps","mapproj","mapdata","MASS","quantreg","rlang","scales","survey","srvyr","viridis","viridisLite","devtools")
install.packages(my_packages)
```

# some demography packages

This is normal text because it's not inside backticks.

Put `R` code inside triple backticks. If you don't want it to actually execute when we generate this document, add the statement
`eval = FALSE` inside the curly brackets. You need the curly brackets anyway to tell the chunk that it's `R`: `{r}`.
```{r, eval = FALSE}
demog_packages <- c("MortalitySmooth", "ungroup", "HMDHFDplus",
					"DemoDecomp", "MortalityLaws")
install.packages(demog_packages)
```

## Forthcoming
I need to come up with a solution to install things from github without using the devtools package. Likely this will involve downloading the packages and installing them locally the old-fashioned way.

## addins

There are some add-ins for Rstudio that can make you life easier, like for tidying your code in certain ways.
```{r, eval = FALSE}
install.packages("addinslist")
library(devtools)
install_github("seasmith/AlignAssign")
```


```{r, eval = FALSE}

# I aligned these using the AlignAssign addin package.
# If you have tons of lines of code, you might want to 
# think about that!
IwriteCode          <- 10
ThisISanotherObject <- 5
```

