---
title: "Class Preparation"
author: "Tim Riffe"
date: "June 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You can safely ignore this file
This is a script I just used for warm-up.

## remake final figure from previous session
```{r, message = FALSE}
library(tidyverse)
library(gapminder)
library(ggplot2)
library(socviz)
library(here)
```

This was the object we were using:
```{r}
by_country <- organdata %>% 
	group_by(consent_law, country) %>%
	# nice trick
	summarize_if(is.numeric, list(mean = mean, sd=sd), na.rm = TRUE) %>%
	ungroup()
```

The penultimate plot we made was a dotplot, where countries are ordered by their mean donor rate and colored by consent law. Recall, all those mappings declared inside the `aes()` function!
```{r}
p <- ggplot(data = by_country,
			mapping = aes(
				x = donors_mean, 
				y = reorder(country, donors_mean),
						  color = consent_law))
p + geom_point(size = 3) 
```
The choice to draw points is declared via the `geom`. Here's the same plot but showing all datapoints (trends not revealed). TO avoid overlap, we jitter.

```{r}
p <- ggplot(data = organdata,
			mapping = aes(x = reorder(country, donors, na.rm =TRUE),
						  y = donors, color = world))
p + geom_jitter(position = position_jitter(width = .15)) + coord_flip() 
```

Rather than color, one could use facetting to separate groups, and this might do a better job of revealing differences in patterns between groups.

```{r}
p <- ggplot(data = by_country,
			mapping = aes(x = donors_mean,
						  y = reorder(country, donors_mean)))
p + geom_point(size = 3) +
	facet_wrap(~consent_law, scales = "free_y", ncol = 1) +
	labs(x = "Donor Procurement Rate", y = "")
```

Here we've reverted back to points for means, but we can also indicate the range with a line segments, a sort of alternative to the boxplots we drew previously. The `geom` for that is `geom_pointrange`, where we explicit about how to draw the upper and lower bounds to the segments. Here +/- one sd.

```{r}
p <- ggplot(data = by_country, mapping = aes(x = reorder(country, donors_mean), y = donors_mean))
p + geom_pointrange(mapping = aes(ymin = donors_mean - donors_sd, ymax = donors_mean + donors_sd)) +
	labs(y = "Donor Procurement Rate", x = "") + coord_flip()
```

How about a labelled scatterplot, in this case showing the relationship between road coverage and donors: better transport (or more accidents?) = more donors? The point here is to use `geom_text()` for adding text to the plot.
```{r}
p <- ggplot(data = by_country, mapping = aes(x = roads_mean, y = donors_mean))
p + geom_point() + geom_text(mapping = aes(label = country))
```

uh oh, a bunch of labels are overlapping both with each other and with the data points. We can solve the second one using `hjust = 1` to right justify it:
```{r}
p <- ggplot(data = by_country, mapping = aes(x = roads_mean, y = donors_mean))
p + geom_point() + geom_text(mapping = aes(label = country), hjust = 1)
```
But there's still overlapping. There's a package just for solving this problem in `ggplot`: `ggrepel`.

```{r}
library(ggrepel)
p <- ggplot(data = by_country, mapping = aes(x = roads_mean, y = donors_mean))
p + geom_point() + geom_text_repel(mapping = aes(label = country), hjust = 1)
```
See, it even draws pointers where necessary. Neat trick, saves you an hour or more of fiddling. Optional- recreate the elections graphic from the book. To label just certain points specify the `data` argument inside `geom_text()` or `geom_text_repel()` where the dataset is cut down to just the needed points using `subset()` or similar.

You can add ad hoc text to plots using the `annotate()` geom. Do this liberally!

```{r}
p <- ggplot(data = organdata, mapping = aes(x = roads, y = donors))
p + geom_point() +
	annotate(geom = "text", 
			 x = 120, y = 33, # explicit location
			 label = "A surprisingly high \nrecovery rates.", hjust = 1)
```
Or you might prefer to demarcate a region:

```{r}
p <- ggplot(data = organdata, mapping = aes(x = roads, y = donors))
p + geom_point() +
	annotate(geom = "rect",
			 xmin = 125, xmax = 155,
			 ymin = 30, ymax = 35, 
			 fill = "red", alpha = 0.2) +
	annotate(geom = "text", 
			 x = 157, y = 33, # explicit location
			 label = "A surprisingly high \nrecovery rate.", hjust = 0)
```

Here is an involved example of `ggrepel` and annotation to make a finely curated scatterplot of US presidencies.

```{r, fig.width = 10, fig.height=10}
elections_historic %>% select(2:7)

p_title    <- "Presidential Elections: Popular & Electoral College Margins"
p_subtitle <- "1824-2016"
p_caption  <- "Data for 2016 are provisional"
x_label    <- "Winner's share of Popular Vote"
y_label    <- "Winner's share of Electoral College Votes"

p <- ggplot(data = elections_historic, 
			mapping = aes(x = popular_pct,
						  y = ec_pct,
						  label = winner_label))
p_out <- p + geom_hline(yintercept = 0.5, size = 1.4, color = gray(.8)) +
	geom_vline(xintercept = 0.5, size = 1.4, color = gray(.8)) + 
	geom_point() +
	geom_text_repel() + 
	scale_x_continuous(labels = scales::percent) + 
	scale_y_continuous(labels = scales::percent) + 
	labs(x = x_label, 
		 y = y_label, 
		 title = p_title, 
		 subtitle = p_subtitle,
		 caption = p_caption)
p_out
# ggsave("Elections.pdf", plot = p_out, height = 10, width = 10)
# getwd()
```

# Chapter 6

```{r}
p <- ggplot(data = gapminder,
			mapping = aes(x = log(gdpPercap),
						  y = lifeExp))

p + geom_point(alpha = .1) + 
	geom_smooth(color = "tomato", fill = "tomato", method = MASS::rlm)+ 
	geom_smooth(color = "steelblue", fill = "steelblue", method = lm)

p + geom_point(alpha = .1) + 
	geom_smooth(color = "tomato", fill = "tomato", method = "lm", size = 1.2, 
				formula = y ~ splines::bs(x,3))

p + geom_point(alpha = .1) + 
	geom_quantile(color = "tomato", size = 1.2, method = "rqss",
				  lambda = 1, quantiles = c(.2,.5,.85))
```










